### 2025-06-06 paper/results.tex
Update results.

Building on the theoretical framework of Partially Observable Active Markov Games developed in the previous section, we now apply our POLARIS algorithm to concrete social learning scenarios. This implementation demonstrates how our approach bridges theoretical economic models with computational reinforcement learning methods, allowing us to validate theoretical predictions while uncovering new insights about strategic adaptation in partial observability settings.

We implement our Partially Observable Active Markov Game framework to model the two canonical social learning scenarios discussed in the background section: strategic experimentation with observable rewards and learning without experimentation through action observation. By reformulating these economic models as POAMGs, we demonstrate how our framework can capture the sophisticated learning dynamics present in these scenarios while enabling the discovery of complex strategies through our POLARIS algorithm. \footnote{Complete Python implementation is available at \url{https://github.com/ecdogaroglu/polaris}.}

@@

Before presenting our experimental results, we must address two fundamental challenges that arise when applying reinforcement learning to social learning environments. These challenges stem from a fundamental mismatch between the strategic requirements of economic theory and the constraints of computational implementation, ultimately necessitating a novel experimental methodology that focuses on within-episode dynamics rather than cross-episode equilibrium convergence.

\subsection{Ex Ante Strategy Optimization vs Ex Post Convergence}

Implementing reinforcement learning in social learning environments reveals a fundamental tension between economic theory and computational practice. In economic social learning models, agents are expected to develop strategies that perform optimally across all possible states of the world—an ex ante perspective where the optimal strategy accounts for uncertainty about which state will be realized. Such strategies require maintaining flexibility and developing sophisticated mixed approaches that can respond appropriately whether the true state turns out to be A or B, good or bad.

However, our computational implementation reveals that agents instead converge to state-specific optimal allocations—an ex post perspective where behavior is optimized for the particular state that has been realized. This distinction creates a critical incompatibility in binary social learning environments we consider here. In strategic experimentation, if the risky arm is revealed to be good, rational agents should allocate all resources to it; conversely, if it performs poorly, optimal behavior requires complete avoidance. In learning without experimentation, once the true state is A, the optimal action is unambiguously action A, while in state B, only action B is rational. These ex post strategies are not merely different from the ex ante approach—they are fundamentally incompatible policy objectives that cannot be reconciled through standard learning approaches.

The computational challenge emerges because reinforcement learning agents, operating within our episodic framework, can only observe and learn from one realized state per episode. This structure forces agents away from developing the sophisticated ex ante strategies characterized in the theoretical models toward state-specific ex post optimization. Instead of learning "if state = A, do X; if state = B, do Y" as economic theory prescribes, agents learn "do X" during episodes initialized with state A, then encounter conflicting objectives to "do Y" during subsequent episodes with state B.

A natural approach to analyzing multi-agent strategic behavior in these environments would be to evaluate reinforcement learning models that are trained across different belief states or environmental configurations. This methodology could theoretically enable us to examine how agents would behave under the full spectrum of possible conditions, potentially revealing the sophisticated mixed strategies and state-contingent policies that characterize optimal ex ante behavior in economic theory. To test this approach, we attempted to recover ex ante policy analysis by inverting a linear grid of possible beliefs into latent belief representations and feeding these directly into trained policy networks. The goal was to evaluate how agents would behave across the full spectrum of possible belief states, potentially revealing the sophisticated mixed strategies characteristic of economic equilibria. However, this approach consistently failed as the trained networks would invariably follow the optimal action corresponding to the last true state they were trained on, regardless of the belief input provided. This confirmed that the learned policies had become fundamentally state-specific rather than developing the belief-contingent strategies required for ex ante optimization, making model evaluation for ex ante analysis impossible due to the underlying technical constraint of catastrophic forgetting.

@@

The ex ante versus ex post discrepancy described above manifests as a severe technical challenge when implementing reinforcement learning in these environments: catastrophic forgetting. Since the behavioral requirements for different state realizations are diametrically opposed, continuous training across episodes with different realized states inevitably leads to catastrophic forgetting. Each episode's learning systematically destroys the value functions and policies acquired in previous episodes with different state realizations, preventing the emergence of stable, discriminative strategies that could respond appropriately to environmental uncertainty as economic theory requires.

To address this fundamental limitation, we explored several established catastrophic forgetting mitigation techniques \citep{vandeven2024continuallearningcatastrophicforgetting}, including Elastic Weight Consolidation \citep{kirkpatrick2017overcoming}, Synaptic Intelligence \citep{zenke2017continuallearningsynapticintelligence}, alternating state presentation schedules, and state-aware replay buffers that integrate experiences from previous environmental configurations. However, these approaches proved ineffective because they assume some degree of transferability or compatibility between learning tasks. When ex post optimal policies require completely opposite actions across states, traditional continual learning techniques cannot meaningfully "consolidate" or "protect" previous learning without fundamentally undermining current learning objectives.

This constraint necessitates resetting agents at the beginning of each episode rather than allowing continuous learning across episodes, which limits our ability to directly evaluate the sophisticated equilibrium strategies described in both \citet{keller2020undiscounted} and related theoretical work. These equilibria require agents to maintain stable ex ante strategies that can discriminate between environmental conditions and respond appropriately—capabilities that prove incompatible with standard reinforcement learning when agents can only experience one state realization per episode. Consequently, our experimental methodology focuses on within-episode learning dynamics and cross-episode pattern analysis to extract meaningful insights about multi-agent strategic behavior despite these fundamental constraints.

@@

\paragraph{Stage 1: Within-Episode Role Identification} For each episode $k$, we identify agents that exhibit extreme performance characteristics. In strategic experimentation, we track cumulative allocation patterns:
\begin{equation}
    i_{\text{high}}^k = \arg\max_i \sum_{t=1}^T a_{t,k}^i, \quad 
    i_{\text{low}}^k = \arg\min_i \sum_{t=1}^T a_{t,k}^i
\end{equation}

In learning without experimentation, we identify the fastest and slowest learning agents by fitting learning rates:
\begin{equation}
    i_{\text{fast}}^k = \arg\max_i r_k^i, \quad 
    i_{\text{slow}}^k = \arg\min_i r_k^i
\end{equation}
where $r_k^i$ is agent $i$'s learning rate in episode $k$.

\paragraph{Stage 2: Cross-Episode Pattern Aggregation} We then aggregate the performance of identified extreme agents across episodes to examine systematic patterns:
\begin{equation}
    \bar{P}_t^{\text{extreme}} = \frac{1}{n_{\text{episodes}}} \sum_{k=1}^{n_{\text{episodes}}} P_{t,k}^{i_{\text{extreme}}^k}
\end{equation}
where $P_{t,k}^i$ represents the relevant performance metric (cumulative allocation, learning accuracy, etc.) for agent $i$ at time $t$ in episode $k$.

This methodology reveals whether observed disparities represent structural features of multi-agent learning or merely artifacts of specific agent characteristics, as different agents assume extreme roles across different episodes.

@@

For each configuration, we run experiments with horizon lengths of 50 steps for comparative sweeps across group sizes, focusing on the emergence of allocation patterns and their relationship to theoretical predictions. Our analysis emphasizes within-episode learning dynamics rather than cross-episode equilibrium convergence due to the catastrophic forgetting constraint.

@@

\paragraph{Strategic Action Convergence} POLARIS agents successfully learn to converge toward state-dependent optimal allocation strategies. In good states, agents increase their allocation to the risky arm over time, while in bad states, they learn to favor the safe alternative. This convergence demonstrates that the discretized action representation effectively captures the continuous allocation dynamics of the theoretical model, enabling agents to discover appropriate exploration-exploitation strategies through reinforcement learning.

\paragraph{Belief Dynamics and Social Learning Effects} Figure \ref{fig:belief_accuracy_over_time} reveals fluctuating belief patterns that indicate the importance of neighbor action observation in the learning process. These fluctuations demonstrate that agents actively incorporate information from neighbors' allocation decisions, leading to more informed belief updating compared to purely individual learning. The variability in belief trajectories reflects the ongoing process of social information aggregation, where agents continuously refine their understanding based on both private signals and observed neighbor behaviors.

\begin{figure}[!htbp]
    \centering
        \includegraphics[width=\textwidth]{../code/polaris/docs/highest_lowest_cumulative_allocators_good_state.png}
        \caption{Highest and lowest cumulative allocators in the good state.}
        \label{fig:highest_lowest_cumulative_allocators_good_state}
\end{figure}

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=\textwidth]{../code/polaris/docs/highest_lowest_cumulative_allocators_bad_state.png}
    \caption{Highest and lowest cumulative allocators in the bad state.}
    \label{fig:highest_lowest_cumulative_allocators_bad_state}
\end{figure}
\paragraph{Dynamic Role Assignment and Allocation Disparities} Figures \ref{fig:highest_lowest_cumulative_allocators_good_state} and \ref{fig:highest_lowest_cumulative_allocators_bad_state} demonstrate a systematic widening of the gap between highest and lowest cumulative allocators as group size increases. This pattern reveals the emergence of dynamic role assignment in multi-agent learning, where agents naturally differentiate into information generators and information exploiters within each episode. In strategic experimentation, this manifests as some agents reducing individual experimentation when benefiting from others' information generation (free-riding effects), while enabling top experimenters to allocate more aggressively when supported by group learning (encouragement effects).

Crucially, our two-staged analysis methodology reveals that these disparities emerge consistently across episodes, with different agents assuming the roles of highest and lowest allocators in different episodes. This dynamic role rotation—parallel to the information generator/exploiter roles observed in learning without experimentation—indicates that the patterns reflect structural features of multi-agent learning rather than fixed agent characteristics. The systematic nature of this role assignment across both experimental scenarios suggests a fundamental principle of social learning: agents naturally specialize into complementary roles that enhance collective information processing, whether through experimentation patterns in strategic contexts or exploration variance in pure learning settings.

@@

We implement comprehensive experiments to validate these theoretical predictions and explore the learning dynamics under various network structures. Our experimental protocol examines network sizes of $n \in \{1, 2, 4, 8\}$ agents; network topologies including complete, ring, star, and random networks; signal accuracy of $p = 0.75$ (we use the same signal accuracy as in the paper's example where $r_{\text{aut}} \approx 0.14$, $r_{\text{crd}} \approx 0.55$, and $r_{\text{bdd}} \approx 1.10$); and learning metrics consisting of empirical learning rates extracted via log-linear regression.

Following our two-staged analysis methodology described in Section \ref{sec:two-staged-analysis}, we track each agent's incorrect action probability over time and estimate individual learning rates by fitting log-linear regressions to identify the fastest and slowest learners within each episode. We then aggregate the performance of these extreme learners across episodes to examine whether the observed learning disparities represent structural features of multi-agent learning rather than artifacts of specific agent characteristics.

Our experimental results provide strong validation of the theoretical predictions while revealing additional insights into the learning dynamics. We begin by addressing a key methodological consideration before examining our main findings.

\paragraph{Methodological Context} The theoretical autarky rate of $r_{\text{aut}} \approx 0.14$ (derived using large deviation theory for random walks) differs substantially from the empirical single-agent POLARIS performance of approximately $0.35$ (Figure \ref{fig:size_comparison}). This difference reflects fundamental methodological distinctions: the theoretical analysis assumes agents follow the maximum likelihood strategy with perfect signal processing, while POLARIS agents learn policies through reinforcement learning with our constructed reward function and Transformer-based processing. Additionally, the theoretical rate derives from asymptotic analysis as $t \to \infty$, whereas our empirical measurements occur over finite horizons. Despite these methodological differences, both approaches provide complementary insights into learning performance bounds and achievable rates in practice.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=\textwidth]{../code/polaris/docs/fastest_slowest_network_sizes_evolution.png}
    \caption{Fastest vs slowest agent learning trajectories across network sizes. Each subplot shows the mean incorrect action probability over time with $95\%$ confidence intervals (shaded regions) for the fastest (green) and slowest (red) learning agents. The learning rates $r$ are calculated using log-linear regression. Results demonstrate the learning barrier effect where slowest agents remain bounded while fastest agents achieve superior performance in larger networks.}
    \label{fig:size_comparison}
\end{figure}

@@

\paragraph{Learning Barriers} Across all network configurations and episodes, we consistently observe a significant reduction in the slowest-learning agent's rate compared to the autarky rate. Figure \ref{fig:size_comparison} demonstrates this barrier effect by aggregating the trajectories of whichever agent learns slowest in each episode. This finding directly supports the theoretical prediction of Theorem 1. While we cannot observe whether the bound significantly exceeds the autarky rate as the theorem suggests—such validation would require much larger networks and higher computational resources—the consistent emergence of learning barriers across all configurations provides strong empirical support for the theoretical framework.

\paragraph{Coordination Benefits} In networks with $n \geq 4$ agents, the fastest-learning agents consistently demonstrate superior performance compared to isolated agents, confirming the coordination benefits predicted by Theorem 2. This demonstrates that social learning enables some agents to surpass single-agent performance, though at the expense of others. The systematic nature of this improvement across different network sizes suggests that the coordination benefits are robust features of multi-agent social learning rather than artifacts of specific configurations. However, similar to our findings on learning barriers, we cannot observe whether the learning rate of the slowest-learning agent can surpass the autarky rate for sufficiently large networks as the theorem suggests.

@@

Crucially, this division of labor emerges dynamically in both scenarios—different agents assume these roles across episodes (Figure \ref{fig:slowest_agent_network_positions}), suggesting the phenomenon arises from learning dynamics rather than fixed agent characteristics. This consistency across fundamentally different social learning environments demonstrates that dynamic role assignment represents a robust feature of multi-agent learning systems, where complementary specialization enhances collective information processing regardless of whether the mechanism operates through allocation patterns or exploration variance.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=\textwidth]{../code/polaris/docs/fastest_slowest_network_types_evolution.png}
    \caption{Fastest vs slowest agent learning trajectories across network topologies. Each subplot compares performance in different network structures (complete, ring, star, random) with 4 agents. The confidence intervals reveal how network topology affects the learning rate disparity between fastest and slowest agents, with complete networks showing the most pronounced differentiation due to enhanced information sharing.}
    \label{fig:type_comparison}
\end{figure}

\subparagraph{Network Structure \& Information Flow} Figure \ref{fig:type_comparison} reveals how network topology influences learning dynamics. Complete networks facilitate the most effective information sharing, producing the largest performance gaps between fastest and slowest learners. Ring networks show more uniform learning rates due to limited information flow, while star networks create pronounced differences between central and peripheral agents. These patterns align with theoretical insights that information aggregation depends critically on network structure, though the specific mechanisms through which topology affects learning warrant further investigation.
\iffalse
To understand the underlying mechanisms determining which agents become information generators, we examine the roles of network position and signal quality—two factors that could potentially explain the emergence of learning disparities.

@@

Crucially, our episode-wise analysis methodology shows that the theoretical bounds from \citet{brandl2024} represent robust properties of the multi-agent learning system rather than characteristics of specific agents. In any given episode, different agents may assume the roles of fastest or slowest learners, yet the aggregate behavior consistently respects the theoretical limits. This emergent division of labor—where some agents naturally become information generators while others exploit this information—arises from the learning dynamics themselves rather than from predetermined agent roles, providing computational validation for the theoretical insights about social learning barriers and coordination benefits despite the equilibrium convergence limitations discussed in our methodological framework.

>>>>

Building on the theoretical framework of Partially Observable Active Markov Games developed in the previous section, we now apply our POLARIS algorithm to concrete social learning scenarios: strategic experimentation and learning without experimentation. This implementation demonstrates how our approach bridges theoretical economic models with computational reinforcement learning methods, allowing us to validate theoretical predictions while uncovering new insights about strategic adaptation in partial observability settings.\footnote{Complete Python implementation is available at \url{https://github.com/ecdogaroglu/polaris}.}

@@

Theoretical frameworks —both in economics and in our POAMG framework— envision agents developing strategies that perform optimally across all possible states of the world. This \emph{ex ante} perspective requires agents to maintain flexibility and develop sophisticated mixed approaches that can respond appropriately regardless of which state is ultimately realized. Agents should learn conditional strategies of the form "if state = A, do X; if state = B, do Y," accounting for uncertainty about the true environmental state.

The computational approach, however, reveals a fundamental departure from this theoretical ideal. Reinforcement learning agents instead converge to state-specific optimal allocations—an \emph{ex post} perspective where behavior is optimized for the particular state that has been realized. This distinction creates a critical incompatibility in the social learning environments we study, particularly due to their fixed-state structure.

To understand why this incompatibility emerges, consider the nature of optimal behavior in social learning environments with binary state spaces. In strategic experimentation, if the risky arm is good, rational agents should allocate all resources to it; conversely, if it performs poorly, optimal behavior requires complete avoidance. Similarly, in learning without experimentation, once the true state is A, the optimal action is unambiguously action A, while in state B, only action B is rational. These ex post strategies represent fundamentally incompatible policy objectives that cannot be reconciled through standard learning approaches.

The fixed-state structure of social learning models compounds this challenge. In each episode, agents can only observe and learn from one realized state, forcing them away from the ex ante strategies that theoretical frameworks prescribe. Instead of developing belief-contingent policies, agents learn "do X" during episodes with state A, then encounter conflicting objectives to "do Y" during subsequent episodes with state B.

One natural approach to address this limitation would be to train agents across different environmental configurations, then evaluate their behavior under the full spectrum of possible conditions. However, this approach consistently fails due to a fundamental constraint of neural network learning. When network parameters are continuously updated across episodes, action-value functions adapt to the most recent state, leading to policies that are only optimal for the last state encountered. This phenomenon—known as \emph{catastrophic forgetting}—fundamentally constrains model evaluation for ex ante analysis.

@@

Catastrophic forgetting represents a fundamental limitation in neural network learning where the acquisition of new knowledge systematically interferes with previously learned information \citep{vandeven2024continuallearningcatastrophicforgetting}. This phenomenon occurs when neural networks, optimized through gradient descent, overwrite previously learned weights and representations as they adapt to new tasks or environments. While various techniques have been developed to mitigate this issue in continual learning scenarios—including regularization methods, memory replay systems, and architectural approaches—these solutions typically assume some degree of compatibility or transferability between learning objectives \citep{kirkpatrick2017overcoming, zenke2017continuallearningsynapticintelligence, rebuffi2017icarl, rusu2016progressive}.

In our social learning environments, this compatibility assumption fails completely. When the optimal policy in state A requires action X while the optimal policy in state B demands the opposite action Y, no meaningful knowledge transfer is possible between episodes with different state realizations. The literature on negative transfer confirms that attempting to force knowledge sharing between fundamentally incompatible tasks degrades performance rather than enhancing it \citep{ahn2024catastrophic, wu2021online}. Traditional continual learning mechanisms that aim to "consolidate" or "protect" previous learning become counterproductive when such protection would directly undermine the acquisition of conflicting but equally necessary knowledge.

These fundamental constraints require us to reset agent parameters at the beginning of each episode rather than allowing continuous learning across state realizations. While this methodological necessity prevents direct measurement of the equilibrium strategies that economic theory and our POAMG framework predict, it does not invalidate our computational approach. Rather, it compels us to develop alternative analytical methodologies that can extract meaningful insights about multi-agent strategic behavior despite these inherent limitations of applying reinforcement learning to environments with conflicting learning objectives.

Our experimental methodology therefore focuses on within-episode learning dynamics and cross-episode pattern analysis. By examining how agents adapt their behavior within episodes and identifying systematic patterns across episodes, we can still characterize the essential features of multi-agent social learning, even when direct equilibrium convergence measurement is precluded by computational limitations.

@@

\paragraph{Stage 1: Within-Episode Role Identification} For each episode $e$, we identify agents that exhibit extreme performance characteristics. In strategic experimentation, we track cumulative allocation patterns:
\begin{equation}
    i_{\text{high}}^e = \arg\max_i \sum_{t=1}^T a_{t,e}^i, \quad 
    i_{\text{low}}^e = \arg\min_i \sum_{t=1}^T a_{t,e}^i
\end{equation}

In learning without experimentation, we identify the fastest and slowest learning agents by fitting learning rates:
\begin{equation}
    i_{\text{fast}}^e = \arg\max_i r_e^i, \quad 
    i_{\text{slow}}^e = \arg\min_i r_e^i
\end{equation}
where $r_e^i$ is agent $i$'s learning rate in episode $e$.

\paragraph{Stage 2: Cross-Episode Pattern Aggregation} We then aggregate the performance of identified extreme agents across episodes to examine systematic patterns:
\begin{equation}
    \bar{P}_t^{\text{extreme}} = \frac{1}{n_{\text{episodes}}} \sum_{e=1}^{n_{\text{episodes}}} P_{t,e}^{i_{\text{extreme}}^e}
\end{equation}
where $P_{t,e}^i$ represents the relevant performance metric (cumulative allocation, learning accuracy, etc.) for agent $i$ at time $t$ in episode $e$.

This methodology reveals whether observed disparities represent structural features of multi-agent learning or merely artifacts of specific agent characteristics like network position, as different agents assume extreme roles across different episodes.

@@

For each configuration, we run experiments for comparative sweeps across group sizes, focusing on the emergence of allocation patterns and their relationship to theoretical predictions. Our analysis emphasizes within-episode learning dynamics rather than cross-episode equilibrium convergence due to the catastrophic forgetting constraint.

@@

\paragraph{Strategic Action Convergence} POLARIS agents successfully learn to converge toward state-dependent optimal allocation strategies. In good states, agents increase their allocation to the risky arm over time, while in bad states, they learn to favor the safe alternative (Figure \ref{fig:allocation_accuracy_over_time}). This convergence demonstrates that the discretized action representation effectively captures the continuous allocation dynamics of the theoretical model, enabling agents to discover appropriate exploration-exploitation strategies through reinforcement learning.

\paragraph{Belief Dynamics and Social Learning Effects} Figure \ref{fig:belief_accuracy_over_time} reveals fluctuating belief patterns alongside stable allocation policies. This demonstrates how POLARIS agents incorporate social information through the inference module while maintaining strategic coherence. The belief variability reflects ongoing social learning as agents observe noisy private signals, yet their policies converge reliably toward optimal state-specific strategies.

\begin{figure}[!htbp]
    \centering
        \includegraphics[width=\textwidth]{../code/polaris/docs/highest_lowest_cumulative_allocators_good_state.png}
        \caption{Highest and lowest cumulative allocators in the good state. (Average over 20 episodes)}
        \label{fig:highest_lowest_cumulative_allocators_good_state}
\end{figure}

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=\textwidth]{../code/polaris/docs/highest_lowest_cumulative_allocators_bad_state.png}
    \caption{Highest and lowest cumulative allocators in the bad state. (Average over 20 episodes)}
    \label{fig:highest_lowest_cumulative_allocators_bad_state}
\end{figure}
\paragraph{Dynamic Role Assignment and Allocation Disparities} Figures \ref{fig:highest_lowest_cumulative_allocators_good_state} and \ref{fig:highest_lowest_cumulative_allocators_bad_state} demonstrate a systematic widening of the gap between highest and lowest cumulative allocators as group size increases. This pattern reveals the emergence of dynamic role assignment in multi-agent learning, where agents naturally differentiate into information generators and information exploiters within each episode. In strategic experimentation, this manifests as some agents reducing individual experimentation when benefiting from others' information generation (free-riding effects), while enabling top experimenters to allocate more aggressively when supported by group learning (encouragement effects). Interestingly, despite the presence of free-riding behavior typically associated with inefficiency in economic literature, our results show that cumulative allocation in larger networks substantially exceeds autarky levels in both good and bad states, suggesting that the social learning benefits and encouragement effects more than compensate for the traditional inefficiencies of free-riding in these multi-agent learning environments.

@@

We implement comprehensive experiments to validate these theoretical predictions and explore the learning dynamics under various network structures. Our experimental protocol examines network sizes of $n \in \{1, 2, 4, 8\}$ agents; network topologies including complete, ring, star, and random networks; signal accuracy of $p = 0.75$ \iffalse(we use the same signal accuracy as in the paper's example where $r_{\text{aut}} \approx 0.14$, $r_{\text{crd}} \approx 0.55$, and $r_{\text{bdd}} \approx 1.10$)\fi; and learning metrics consisting of empirical learning rates extracted via log-linear regression.

Following our two-staged analysis methodology described in Section \ref{sec:two-staged-analysis}, we track each agent's incorrect action probability as the probability assignment of the policy network to the incorrect action over time and estimate individual learning rates by fitting log-linear regressions to identify the fastest and slowest learners within each episode. We then aggregate the performance of these extreme learners across episodes to examine whether the observed learning disparities represent structural features of multi-agent learning rather than artifacts of specific agent characteristics.

\subsection{Results and Analysis}
\iffalse
Our experimental results provide strong validation of the theoretical predictions while revealing additional insights into the learning dynamics. We begin by addressing a key methodological consideration before examining our main findings.

\paragraph{Methodological Context} The theoretical autarky rate of $r_{\text{aut}} \approx 0.14$ (derived using large deviation theory for random walks) differs substantially from the empirical single-agent POLARIS performance of approximately $0.35$ (Figure \ref{fig:size_comparison}). This difference reflects fundamental methodological distinctions: the theoretical analysis assumes agents follow the maximum likelihood strategy with perfect signal processing, while POLARIS agents learn policies through reinforcement learning with our constructed reward function and Transformer-based processing. Additionally, the theoretical rate derives from asymptotic analysis as $t \to \infty$, whereas our empirical measurements occur over finite horizons. Despite these methodological differences, both approaches provide complementary insights into learning performance bounds and achievable rates in practice.
\fi

@@

\paragraph{Learning Barriers} Across all network configurations and episodes, we consistently observe a significant reduction in the slowest-learning agent's rate compared to the autarky rate. Figure \ref{fig:size_comparison} demonstrates this barrier effect by aggregating the trajectories of whichever agent learns slowest in each episode. This finding directly supports the theoretical prediction of the paper. While we cannot observe whether the bound significantly exceeds the autarky rate as the theorem suggests—such validation would require much larger networks and higher computational resources—the consistent emergence of learning barriers across all configurations provides strong empirical support for the theoretical framework.

\paragraph{Coordination Benefits} In networks with $n \geq 4$ agents, the fastest-learning agents consistently demonstrate superior performance compared to isolated agents, confirming the coordination benefits predicted by the paper. This demonstrates that social learning enables some agents to surpass single-agent performance, though at the expense of others. The systematic nature of this improvement across different network sizes suggests that the coordination benefits are robust features of multi-agent social learning rather than artifacts of specific configurations. However, similar to our findings on learning barriers, we cannot observe whether the learning rate of the slowest-learning agent can surpass the autarky rate for sufficiently large networks as the theorem suggests.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=\textwidth]{../code/polaris/docs/fastest_slowest_network_sizes_evolution.png}
    \caption{Fastest vs slowest agent learning trajectories across network sizes.}
    \label{fig:size_comparison}
\end{figure}

@@

\iffalse
Crucially, this division of labor emerges dynamically in both scenarios—different agents assume these roles across episodes (Figure \ref{fig:slowest_agent_network_positions}), suggesting the phenomenon arises from learning dynamics rather than fixed agent characteristics. This consistency across fundamentally different social learning environments demonstrates that dynamic role assignment represents a robust feature of multi-agent learning systems, where complementary specialization enhances collective information processing regardless of whether the mechanism operates through allocation patterns or exploration variance.
\fi

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=\textwidth]{../code/polaris/docs/fastest_slowest_network_types_evolution.png}
    \caption{Fastest vs slowest agent learning trajectories across network topologies.}
    \label{fig:type_comparison}
\end{figure}

\paragraph{Network Structure \& Information Flow} Figure \ref{fig:type_comparison} reveals how network topology influences learning dynamics. Complete networks facilitate the most effective information sharing, producing the largest performance gaps between fastest and slowest learners. Ring networks show more uniform learning rates due to limited information flow, while star networks create pronounced differences between central and peripheral agents. These patterns align with theoretical insights that information aggregation depends critically on network structure, though the specific mechanisms through which topology affects learning warrant further investigation.
\iffalse
To understand the underlying mechanisms determining which agents become information generators, we examine the roles of network position and signal quality—two factors that could potentially explain the emergence of learning disparities.

@@

Our computational analysis reveals a striking consistency in the emergence of dynamic role assignment across fundamentally different social learning environments. In both strategic experimentation and learning without experimentation, agents naturally differentiate into complementary roles that enhance collective information processing, though the specific mechanisms differ substantially. In strategic experimentation, this differentiation manifests through allocation patterns, where some agents assume higher-risk experimental roles by allocating more resources to uncertain alternatives, while others free-ride on this information generation. In learning without experimentation, the same underlying phenomenon emerges through learning rate disparities and exploration variance, where slower-learning agents exhibit more exploratory behavior that benefits faster-learning agents who can exploit this information revelation. Crucially, these roles are not fixed by agent characteristics or network positions but emerge dynamically through the learning process itself, with different agents assuming information generation and exploitation roles across episodes. Remarkably, in both environments, the presence of free-riding behavior does not lead to the inefficiencies typically predicted by economic theory—cumulative allocations in strategic experimentation and learning rates in learning without experimentation both exceed autarky levels in larger networks, demonstrating that for some agents, social learning benefits and encouragement effects consistently outweigh traditional free-riding costs. This robustness across distinct social learning mechanisms suggests that dynamic role assignment represents a fundamental organizing principle of multi-agent learning systems, enabling some agents to efficiently aggregate information through spontaneous specialization in exploratory behavior.
